Incremental λ-Calculus in Cache-Transfer Style

### 2020-01-21 popl18/cts-relwork.tex
Add lhs2TeX output for archival

\section{Incrementalization case studies}
\label{sec:case-studies}
\label{sec:cts-case-studies}

In this section, we investigate whether our transformations incrementalize
efficiently programs in a typed language such as Haskell.
And indeed, after providing support for the needed bulk operations on sequences,
bags and maps, we successfully transform a few case studies and obtain efficient
incremental programs, that is ones for which incremental computation is faster
than from scratch recomputation.

We type CTS programs by by associating to each function \ensuremath{\Varid{f}} a cache type \ensuremath{\Conid{FC}}.
We can transform programs that use higher-order functions by making closure
construction explicit.

We illustrate those points on three case studies: average computation over bags of
integers, a nested loop over two sequences and a more involved example inspired
by \citeauthor{Koch14}'s work on incrementalizing database queries.

In all cases, we confirm that results are consistent between from scratch
recomputation and incremental evaluation.

Our benchmarks were compiled by GHC 8.0.2. They were run on a 2.60GHz dual core
Intel i7-6600U CPU with 12GB of RAM running Ubuntu 16.04.

\subsection{Averaging integers bags}
\label{sec:incr-an-aver}

Section \cref{sec:motivating-example} motivates our transformation with a
running example of computing the average over a bag of integers.
We represent bags as maps from elements to (possibly negative) multiplicities.
Earlier work \citep{CaiEtAl2014ILC,Koch14} represents bag changes as bags
of removed and added elements: to update element \ensuremath{\Varid{a}} with change \ensuremath{\Varid{da}}, one removes
\ensuremath{\Varid{a}} and adds \ensuremath{\Varid{a}\oplus \Varid{da}}. Instead, our bag changes contain element changes: a
valid bag change is either a replacement change (with a new bag) or a list of atomic
changes.
An atomic change contains an element \ensuremath{\Varid{a}}, a valid change for that element
\ensuremath{\Varid{da}}, a multiplicity \ensuremath{\Varid{n}} and a change to that multiplicity \ensuremath{\Varid{dn}}. Updating a bag
with an atomic change means removing \ensuremath{\Varid{n}} occurrences of \ensuremath{\Varid{a}} and
inserting \ensuremath{\Varid{n}\oplus \Varid{dn}} occurrences of updated element \ensuremath{\Varid{a}\oplus \Varid{da}}.
% Incremental functions can handle such changes
% In other representations, modifying an element
% when representing bag changes as
% bags of additions and removals~\citep{CaiEtAl2014ILC,Koch14}, element ch
% When represe
% In contrast to
% Compared to a representation of bag changes as bags as in
% \citet{CaiEtAl2014ILC} this representation allows us to represent changes to
% elements more directly making some programs more incremental.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Bag}\;\Varid{a}\mathrel{=}\Conid{Map}\;\Varid{a}\;\mathbb{Z}{}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Delta (\Conid{Bag}\;\Varid{a})\mathrel{=}\Conid{Replace}\;(\Conid{Bag}\;\Varid{a})\mid \Conid{Ch}\;[\mskip1.5mu \Conid{AtomicChange}\;\Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{AtomicChange}\;\Varid{a}\mathrel{=}\Conid{AtomicChange}\;\Varid{a}\;(\Delta \Varid{a})\;\mathbb{Z}\;(\Delta \mathbb{Z}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This change type enables both users and derivatives to express efficiently various bag
changes, such as inserting a single element or changing a single element, but
also changing some but not all occurrences of an element in a bag.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{insertSingle}\mathrel{:\mkern-1mu:}\Varid{a}\to \Delta (\Conid{Bag}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\Varid{insertSingle}\;\Varid{a}\mathrel{=}\Conid{Ch}\;[\mskip1.5mu \Conid{AtomicChange}\;\Varid{a}\;\NilC{\Varid{a}}\;\mathrm{0}\;(\Conid{Ch}\;\mathrm{1})\mskip1.5mu]{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{changeSingle}\mathrel{:\mkern-1mu:}\Varid{a}\to \Delta \Varid{a}\to \Delta (\Conid{Bag}\;\Varid{a}){}\<[E]%
\\
\>[B]{}\Varid{changeSingle}\;\Varid{a}\;\Varid{da}\mathrel{=}\Conid{Ch}\;[\mskip1.5mu \Conid{AtomicChange}\;\Varid{a}\;\Varid{da}\;\mathrm{1}\;\NilC{\mathrm{1}}\mskip1.5mu]{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Based on this change structure, we provide efficient primitives and their
derivatives. The CTS variant of \ensuremath{\Varid{map}}, that we call \ensuremath{\Varid{mapC}} takes a function
\ensuremath{\Varid{fC}} in CTS and a bag \ensuremath{\Varid{as}} and produces a bag and a cache. Because \ensuremath{\Varid{map}} is not
self-maintainable the cache stores the arguments \ensuremath{\Varid{fC}} and \ensuremath{\Varid{as}}. In addition the
cache stores for each invocation of \ensuremath{\Varid{fC}}, and therefore for each distinct element
in \ensuremath{\Varid{as}}, the result of \ensuremath{\Varid{fC}} of type \ensuremath{\Varid{b}} and the cache of type \ensuremath{\Varid{c}}.
The incremental function \ensuremath{\Varid{dmapC}} has to produce an
updated cache. We check that this is the same cache that \ensuremath{\Varid{mapC}} would produce
when applied to updated inputs using the QuickCheck library.

Following ideas inspired by~\citet{Rossberg2010fing}, all higher-order functions
(and typically, also their caches) are parametric over cache types of their
function arguments. Here, functions \ensuremath{\Varid{mapC}} and \ensuremath{\Varid{dmapC}} and cache type \ensuremath{\Conid{MapC}}
are parametric over the cache type \ensuremath{\Varid{c}} of \ensuremath{\Varid{fC}} and \ensuremath{\Varid{dfC}}.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\Varid{map}\mathrel{:\mkern-1mu:}(\Varid{a}\to \Varid{b})\to \Conid{Bag}\;\Varid{a}\to \Conid{Bag}\;\Varid{b}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\mathbf{type}\;\Conid{MapC}\;\Varid{a}\;\Varid{b}\;\Varid{c}\mathrel{=}(\Varid{a}\to (\Varid{b},\Varid{c}),\Conid{Bag}\;\Varid{a},\Conid{Map}\;\Varid{a}\;(\Varid{b},\Varid{c})){}\<[E]%
\\
\>[B]{}\Varid{mapC}\mathrel{:\mkern-1mu:}(\Varid{a}\to (\Varid{b},\Varid{c}))\to \Conid{Bag}\;\Varid{a}\to (\Conid{Bag}\;\Varid{b},\Conid{MapC}\;\Varid{a}\;\Varid{b}\;\Varid{c}){}\<[E]%
\\
\>[B]{}\Varid{dmapC}\mathrel{:\mkern-1mu:}(\Delta \Varid{a}\to \Varid{c}\to (\Delta \Varid{b},\Varid{c}))\to \Delta (\Conid{Bag}\;\Varid{a})\to {}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Conid{MapC}\;\Varid{a}\;\Varid{b}\;\Varid{c}\to (\Delta (\Conid{Bag}\;\Varid{b}),\Conid{MapC}\;\Varid{a}\;\Varid{b}\;\Varid{c}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We wrote the \ensuremath{\Varid{length}} and \ensuremath{\Varid{sum}} function used in our benchmarks in terms of
primitives \ensuremath{\Varid{map}} and \ensuremath{\Varid{foldGroup}}. We applied our transformations to get
\ensuremath{\Varid{lengthC}}, \ensuremath{\Varid{dlengthC}}, \ensuremath{\Varid{sumC}} and \ensuremath{\Varid{dsumC}}. This transformation could in
principle be automated. Implementing the CTS variant and CTS derivative of
primitives efficiently requires manual work.

We evaluate whether we can produce an updated result with \ensuremath{\Varid{davgC}} faster than
by from scratch recomputation with \ensuremath{\Varid{avg}}. We use the \ensuremath{\Varid{criterion}} library for
benchmarking. The benchmarking code and our raw results are available in the
supplementarty material. We fix an initial bag of size \ensuremath{\Varid{n}}. It contains the
integers
from \ensuremath{\mathrm{1}} to \ensuremath{\Varid{n}}. We define a sequence of consecutive changes of length \ensuremath{\Varid{r}}
as the insertion of \ensuremath{\mathrm{1}}, the deletion of \ensuremath{\mathrm{1}}, the insertion of \ensuremath{\mathrm{2}}, the deletion
of \ensuremath{\mathrm{2}} and so on. We update the initial bag with these changes, one after
another, to obtain a sequence of input bags.

We measure the time taken by from scratch recomputation. We apply \ensuremath{\Varid{avg}} to
each input bag in the sequence and fully force all results. We report the time
from scratch recomputation takes as this measured time divided by the number \ensuremath{\Varid{r}}
of input bags. We measure the time taken by our CTS variant \ensuremath{\Varid{avgC}} similarly.
We make sure to fully force all results and caches \ensuremath{\Varid{avgC}} produces on the sequence
of input bags.

We measure the time taken by our CTS derivative \ensuremath{\Varid{davgC}}. We assume a fully
forced initial result and cache. We apply \ensuremath{\Varid{davgC}} to each change in the
sequence of bag changes using the cache from the previous invocation. We then
apply the result change to the previous result. We fully force the obtained
sequence of results. Finally we measure the time taken to only produce the sequence
of result changes without applying them.

Because of laziness, choosing what exactly to measure is tricky.
We ensure that reported times include the entire cost of computing the whole
sequence of results. To measure this cost, we ensure any thunks within the
sequence are forced. We need not force any partial results, such as output
changes or caches. Doing so would distort the results because forcing the whole
cache can cost asymptotically more than running derivatives. However, not using
the cache at all underestimates the cost of incremental computation. Hence, we
measure a sequence of consecutive updates, each using the cache produced by the
previous one.

\begin{figure}
  \footnotesize
  \begin{subfigure}[l]{0.495\linewidth}
  \begin{tikzpicture}
    \begin{axis}[
        width=\linewidth,
        height=\linewidth,
        xlabel=input size,
        ylabel=time in milliseconds,
        y filter/.code={\pgfmathparse{#1*1000}\pgfmathresult},
        legend pos=north west
      ]
      \addplot table[x=size, y=from_scratch, col sep=comma]{\poplPath{benchmark-results/average.csv}};
      \addlegendentry{original}
      \addplot table[x=size, y=caching, col sep=comma]{\poplPath{benchmark-results/average.csv}};
      \addlegendentry{caching}
      \addplot table[x=size, y=change_only, col sep=comma]{\poplPath{benchmark-results/average.csv}};
      \addlegendentry{change only}
      \addplot table[x=size, y=incremental, col sep=comma]{\poplPath{benchmark-results/average.csv}};
      \addlegendentry{incremental}
    \end{axis}
  \end{tikzpicture}
  \caption{Benchmark results for \ensuremath{\Varid{avg}}}
  \label{fig:average-results}
\end{subfigure}
\hfill
\begin{subfigure}[r]{0.495\linewidth}
  \begin{tikzpicture}
    \begin{axis}[
        width=\linewidth,
        height=\linewidth,
        xlabel=input size,
        ylabel=time in milliseconds,
        y filter/.code={\pgfmathparse{#1*1000}\pgfmathresult},
        legend pos=north west
      ]
      \addplot table[x=size, y=from_scratch, col sep=comma]{\poplPath{benchmark-results/indexedJoin.csv}};
      \addlegendentry{original}
      \addplot table[x=size, y=caching, col sep=comma]{\poplPath{benchmark-results/indexedJoin.csv}};
      \addlegendentry{caching}
      \addplot table[x=size, y=change_only, col sep=comma]{\poplPath{benchmark-results/indexedJoin.csv}};
      \addlegendentry{change only}
      \addplot table[x=size, y=incremental, col sep=comma]{\poplPath{benchmark-results/indexedJoin.csv}};
      \addlegendentry{incremental}
    \end{axis}
  \end{tikzpicture}
  \caption{Benchmark results for \ensuremath{\Varid{totalPrice}}}
  \label{fig:indexedJoin-results}
\end{subfigure}
\end{figure}

The plot in \cref{fig:average-results} shows execution time versus the size
\ensuremath{\Varid{n}} of the initial input. To produce the initial result and cache, \ensuremath{\Varid{avgC}} takes
longer than the original \ensuremath{\Varid{avg}} function takes to produce just the result. This
is to be expected. Producing the result incrementally is much faster than from
scratch recomputation. This speedup increases with the size of the initial bag.
For an initial bag size of 800 incrementally updating the result is 70 times faster
than from scratch recomputation.

The difference between only producing the output change versus producing
the output change and updating the result is very small in this example. This is
to be expected because in this example the result is an integer and updating
and evaluating an integer is fast. We will see an example where there is a
bigger difference between the two.

\subsection{Nested loops over two sequences}
\label{sec:incr-nest-loop}

We implemented incremental sequences and related primitives following \citet{Firsov2016purely}: our
change operations and first-order operations (such as \ensuremath{\Varid{concat}}) reuse their
implementation. On the other hand, we must extend higher-order operations such
as \ensuremath{\Varid{map}} to handle non-nil function changes and caching. A correct and efficient
CTS incremental \ensuremath{\Varid{dmapC}} function has to work differently depending on whether the
given function change is nil or not: For a non-nil function change it has to
go over the input sequence; for a nil function change it can avoid that.

\ps{Briefly describe change structure for sequences}

Consider the running example again, this time in A'NF\@. The partial application of
a lambda lifted function constructs a closure. We made that explicit with a
\ensuremath{\Varid{closure}} function.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\Varid{cartesianProduct}\;\Varid{xs}\;\Varid{ys}\mathrel{=}\mathbf{let}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{mapPairYs}\mathrel{=}\Varid{closure}\;\Varid{mapPair}\;\Varid{ys}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{xys}\mathrel{=}\Varid{concatMap}\;\Varid{mapPairYs}\;\Varid{xs}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\mathbf{in}\;\Varid{xys}{}\<[E]%
\\[\blanklineskip]%
\>[3]{}\Varid{mapPair}\;\Varid{ys}\mathrel{=}\lambda \Varid{x}\to \mathbf{let}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{pairX}\mathrel{=}\Varid{closure}\;\Varid{pair}\;\Varid{x}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Varid{xys}\mathrel{=}\Varid{map}\;\Varid{pairX}\;\Varid{ys}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\mathbf{in}\;\Varid{xys}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

While the only valid change for closed functions is their nil change, for closures
we can have non-nil function changes. We represent closed functions and closures
as variants of the same type. Correspondingly we represent changes to a closed
function and changes to a closure as variants of the same type of function changes.
We inspect this representation at runtime to find out if a function change
is a nil change.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathbf{data}\;\Conid{Fun}\;\Varid{a}\;\Varid{b}\;\Varid{c}\;\mathbf{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Closed}\mathrel{:\mkern-1mu:}(\Varid{a}\to (\Varid{b},\Varid{c}))\to \Conid{Fun}\;\Varid{a}\;\Varid{b}\;\Varid{c}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{Closure}\mathrel{:\mkern-1mu:}(\Varid{e}\to \Varid{a}\to (\Varid{b},\Varid{c}))\to \Varid{e}\to \Conid{Fun}\;\Varid{a}\;\Varid{b}\;\Varid{c}{}\<[E]%
\\[\blanklineskip]%
\>[3]{}\mathbf{data}\;\Delta (\Conid{Fun}\;\Varid{a}\;\Varid{b}\;\Varid{c})\;\mathbf{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{DClosed}\mathrel{:\mkern-1mu:}(\Delta \Varid{a}\to \Varid{c}\to (\Delta \Varid{b},\Varid{c}))\to \Delta (\Conid{Fun}\;\Varid{a}\;\Varid{b}\;\Varid{c}){}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Conid{DClosure}\mathrel{:\mkern-1mu:}(\Delta \Varid{e}\to \Delta \Varid{a}\to \Varid{c}\to (\Delta \Varid{b},\Varid{c}))\to \Delta \Varid{e}\to \Delta (\Conid{Fun}\;\Varid{a}\;\Varid{b}\;\Varid{c}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

We have also evaluated another representation of function changes with different
tradeoffs where we use defunctionalization instead of closure conversion.
\begin{poplForThesis}
We discuss the use of defunctionalization in \cref{ch:defunc-fun-changes}.
\end{poplForThesis}

We use the same benchmark setup as in the benchmark for the average computation on bags.
The input for size \ensuremath{\Varid{n}} is a pair of sequences \ensuremath{(\Varid{xs},\Varid{ys})}. Each sequence
initially contains the integers from \ensuremath{\mathrm{1}} to \ensuremath{\Varid{n}}. Updating the result in reaction
to a change \ensuremath{\Varid{dxs}} to \ensuremath{\Varid{xs}} takes less time than updating the result in reaction
to a change \ensuremath{\Varid{dys}} to \ensuremath{\Varid{ys}}. The reason is that when \ensuremath{\Varid{dys}} is a nil change we
dynamically detect that the function change passed to \ensuremath{\Varid{dconcatMapc}} is the nil
change and avoid looping over \ensuremath{\Varid{xs}} entirely.

We benchmark changes to the outer sequence \ensuremath{\Varid{xs}} and the inner sequence \ensuremath{\Varid{ys}}
separately. In both cases the sequence of changes are insertions and deletions of
different numbers at different positions.
When we benchmark changes to the outer sequence \ensuremath{\Varid{xs}}
we take this sequence of changes for \ensuremath{\Varid{xs}} and all changes to \ensuremath{\Varid{ys}} will be nil
changes. When we benchmark changes to the inner sequence \ensuremath{\Varid{ys}} we take this sequence
of changes for \ensuremath{\Varid{ys}} and all changes to \ensuremath{\Varid{xs}} will be nil changes.

\begin{figure}
  \footnotesize
  \begin{subfigure}[l]{0.495\linewidth}
  \begin{tikzpicture}
    \begin{axis}[
        width=\linewidth,
        height=\linewidth,
        xlabel=input size,
        ylabel=time in seconds,
        legend pos=north west
      ]
      \addplot table[x=size, y=from_scratch, col sep=comma]{\poplPath{benchmark-results/nestedLoopInner.csv}};
      \addlegendentry{original}
      \addplot table[x=size, y=caching, col sep=comma]{\poplPath{benchmark-results/nestedLoopInner.csv}};
      \addlegendentry{caching}
      \addplot table[x=size, y=change_only, col sep=comma]{\poplPath{benchmark-results/nestedLoopInner.csv}};
      \addlegendentry{change only}
      \addplot table[x=size, y=incremental, col sep=comma]{\poplPath{benchmark-results/nestedLoopInner.csv}};
      \addlegendentry{incremental}
    \end{axis}
  \end{tikzpicture}
  \caption{Benchmark results for cartesian product changing \emph{inner} sequence.}
  \label{fig:nestedLoopInner-results}
\end{subfigure}
\hfill
\begin{subfigure}[r]{0.495\linewidth}
  \begin{tikzpicture}
    \begin{axis}[
        width=\linewidth,
        height=\linewidth,
        xlabel=input size,
        ylabel=time in seconds,
        legend pos=north west
      ]
      \addplot table[x=size, y=from_scratch, col sep=comma]{\poplPath{benchmark-results/nestedLoopOuter.csv}};
      \addlegendentry{original}
      \addplot table[x=size, y=caching, col sep=comma]{\poplPath{benchmark-results/nestedLoopOuter.csv}};
      \addlegendentry{caching}
      \addplot table[x=size, y=change_only, col sep=comma]{\poplPath{benchmark-results/nestedLoopOuter.csv}};
      \addlegendentry{change only}
      \addplot table[x=size, y=incremental, col sep=comma]{\poplPath{benchmark-results/nestedLoopOuter.csv}};
      \addlegendentry{incremental}
    \end{axis}
  \end{tikzpicture}
  \caption{Benchmark results for Cartesian product changing \emph{outer} sequence.}
  \label{fig:nestedLoopOuter-results}
\end{subfigure}
\end{figure}

In this example preparing the cache takes much longer than from scratch
recomputation. The speedup of incremental computation over from scratch
recomputation increases with the size of the initial sequences. It reaches
\ensuremath{\mathrm{2.5}} for a change to the inner sequences and \ensuremath{\mathrm{8.8}} for a change to the outer
sequence when the initial sequences have size \ensuremath{\mathrm{800}}. Producing and fully forcing
only the changes to the results is \ensuremath{\mathrm{5}} times faster for a change to the inner
sequence and \ensuremath{\mathrm{370}} times faster for a change to the outer sequence.

While we do get speedups for both kinds of
changes (to the inner and to the outer sequence) the speedups for changes to the
outer sequence are bigger.
Due to our benchmark methodology we have to fully force updated results. This alone
takes time quadratic in the size of the input sequences \ensuremath{\Varid{n}}. Threrefore we also
report the time it takes to produce and fully force only the output changes which
is of a lower time complexity.

\subsection{Indexed joins of two bags}
\label{sec:incr-an-index}

As expected, we found that we can compose functions into larger and more complex programs and
apply CTS differentiation to get a fast incremental program.

For this example imagine we have a bag of orders and a bag of line items. An
order is a pair of an integer key and an exchange rate represented as an integer.
A line item is a pair of a key of a corresponding order and a price represented
as an integer. The price of an order is the sum of the prices of the
corresponding line items multiplied by the exchange rate. We want to find the
total price defined as the sum of the prices of all orders.

To do so efficiently we build an index for line items and an index for orders.
We represent indexes as maps. We build a map from order key to the sum of the
prices of corresponding line items. Because orders are not necessarily unique
by key and because multiplication distributes over addition we can build an index
mapping each order key to the sum of all exchange rates for this order key.
We then compute the total price by merging the two maps by key, multiplying
corresponding sums of exchange rates and sums of prices. We compute the total
price as the sum of those products.

Because our indexes are maps we need a change structure for maps. We generalize
the change structure for bags by generalizing from multiplicities to arbitrary
values as long as they have a group structure. A change to a map is either a
replacement change (with a new map) or a list of atomic changes. An atomic
change is a key \ensuremath{\Varid{k}}, a valid change to that key \ensuremath{\Varid{dk}}, a value \ensuremath{\Varid{a}} and a valid
change to that value \ensuremath{\Varid{da}}. To update a map with an atomic change means to use
the group structure of the type of \ensuremath{\Varid{a}} to subtract \ensuremath{\Varid{a}} at key \ensuremath{\Varid{k}} and to add \ensuremath{\Varid{a}\oplus \Varid{da}} at key \ensuremath{\Varid{k}\oplus \Varid{dk}}.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Delta (\Conid{Map}\;\Varid{k}\;\Varid{a})\mathrel{=}\Conid{Replace}\;(\Conid{Map}\;\Varid{k}\;\Varid{a})\mid \Conid{Ch}\;[\mskip1.5mu \Conid{AtomicChange}\;\Varid{k}\;\Varid{a}\mskip1.5mu]{}\<[E]%
\\
\>[B]{}\mathbf{data}\;\Conid{AtomicChange}\;\Varid{k}\;\Varid{a}\mathrel{=}\Conid{AtomicChange}\;\Varid{k}\;(\Delta \Varid{k})\;\Varid{a}\;(\Delta \Varid{a}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Bags have a group structure as long as the elements have a group structure. Maps
have a group structure as long as the values have a group structure. This means
for example that we can efficiently incrementalize programs on maps from keys
to bags of elements. We implemented efficient caching and incremental primitives
for maps and verified their correctness with QuickCheck.

To build the indexes, we use a \ensuremath{\Varid{groupBy}} function built from primitive
functions \ensuremath{\Varid{foldMapGroup}} on bags and \ensuremath{\Varid{singleton}} for bags
and maps respectively. While computing the indexes with \ensuremath{\Varid{groupBy}} is
self-maintainable, merging them is not. We need to cache and incrementally
update the intermediately created indexes to avoid recomputing them.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathbf{type}\;\Conid{Order}\mathrel{=}(\mathbb{Z},\mathbb{Z}){}\<[E]%
\\
\>[B]{}\mathbf{type}\;\Conid{LineItem}\mathrel{=}(\mathbb{Z},\mathbb{Z}){}\<[E]%
\\[\blanklineskip]%
\>[B]{}\Varid{totalPrice}\mathrel{:\mkern-1mu:}\Conid{Bag}\;\Conid{Order}\to \Conid{Bag}\;\Conid{LineItem}\to \mathbb{Z}{}\<[E]%
\\
\>[B]{}\Varid{totalPrice}\;\Varid{orders}\;\Varid{lineItems}\mathrel{=}\mathbf{let}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{orderIndex}\mathrel{=}\Varid{groupBy}\;\Varid{fst}\;\Varid{orders}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{orderSumIndex}\mathrel{=}\Varid{\Conid{Map}.map}\;(\Varid{\Conid{Bag}.foldMapGroup}\;\Varid{snd})\;\Varid{orderIndex}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{lineItemIndex}\mathrel{=}\Varid{groupBy}\;\Varid{fst}\;\Varid{lineItems}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{lineItemSumIndex}\mathrel{=}\Varid{\Conid{Map}.map}\;(\Varid{\Conid{Bag}.foldMapGroup}\;\Varid{snd})\;\Varid{lineItemIndex}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{merged}\mathrel{=}\Varid{\Conid{Map}.merge}\;\Varid{orderSumIndex}\;\Varid{lineItemSumIndex}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Varid{total}\mathrel{=}\Varid{\Conid{Map}.foldMapGroup}\;\Varid{multiply}\;\Varid{merged}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\mathbf{in}\;\Varid{total}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks

This example is inspired by \citet{Koch14}.
%
Unlike them, we don't automate indexing, so to get good performance we start
from a program that explicitly uses indexes.
% While indexing arise automatically
% from their transformations,
% to get good performance we must start from a program
% that explicitly uses indexes

We evaluate the performace in the same way we did for the average computation on bags. The
initial input of size \ensuremath{\Varid{n}} is a pair of bags where both contain the pairs
\ensuremath{(\Varid{i},\Varid{i})} for \ensuremath{\Varid{i}} between \ensuremath{\mathrm{1}} and \ensuremath{\Varid{n}}. The sequence of changes are alternations
between insertion and deletion of pairs \ensuremath{(\Varid{j},\Varid{j})} for different \ensuremath{\Varid{j}}. We
alternate the insertions and deletions between the orders bag and the line items
bag.

Our CTS derivative of the original program produces updated results much faster
than from scratch recomputation and again the speedup increases with the size
of the initial bags. For an initial size of the two bags of \ensuremath{\mathrm{800}}
incrementally updating the result is 180 times faster than from scratch
recomputation.

\iftoggle{poplForThesis}{\section}{\subsection}{Limitations and future work}
\iftoggle{poplForThesis}{\let \limSection = \subsection}{\let \limSection = \subsubsection}
\label{sec:cts-limitations}
% Knowing myself, I'm probably doing this too much.
In this section we describe limitations to be addressed in future work.

\pg{Commenting this out, don't believe it any more in general. Unless we
  do a group fold, we need to run \ensuremath{\ominus }!}
% \subsubsection{Inductive data types and structural recursion}

% We believe that we can extend our transformation to work with folds over inductive
% data types.
% Derivatives for folds still need be written by hand, but mainly to handle
% structural insertions or deletions; value changes can be handled by the main
% transformation.
% General recursive functions induce recursive cache types.

% DOES THIS MEAN WE NEVER NEED ANY MAINTENANCE? Probably not. But where do we fail?

>>>>

\section{Related work}
\label{sec:cts-rw}
Of all research on incremental computation in both programming languages and
databases~\citep{Gupta99MMV,Ramalingam93}, we discuss the most closely related works.
\begin{poplForThesis}
Other related work, more closely to cache-transfer style, is
discussed in \cref{sec:rw}.
\end{poplForThesis}

\paragraph{Previous work on cache-transfer-style}
\citet{Liu00}'s work has been the fundamental inspiration to this work, but
her approach has no correctness proof and is restricted to a first-order untyped
language (in part because no absence analysis for higher-order languages was
available). Moreover, while the idea of cache-transfer-style is similar, it's
unclear if her approach to incrementalization would extend to higher-order
programs.
% while her approach to absence analysis was based on available
% technology that did not extend to higher-order programs unlike
% \citet{Sergey2014modular}'s.

\citet{Firsov2016purely} also approach incrementalization by code
transformation, but their approach does not deal with changes to functions.
Instead of transforming functions written in terms of primitives, they provide
combinators to write CTS functions and derivatives together. On the other hand,
they extend their approach to support mutable caches, while restricting to
immutable ones as we do might lead to a logarithmic slowdown.

\paragraph{Finite differencing}
Incremental computation on collections or databases by finite differencing has a long
tradition~\citep{Paige82FDC,Blakeley:1986:EUM}. The most recent and
impressive line of work is the one on DBToaster~\citep{Koch10IQE,Koch14}, which
is a highly efficient approach to incrementalize queries over bags by combining
iterated finite differencing with other program transformations. They show
asymptotic speedups both in theory and through experimental evaluations.
Changes are only allowed for datatypes that form groups (such as bags or certain
maps), but not for instance for lists or sets. Similar ideas were recently
extended to higher-order and nested computation~\citep{Koch2016incremental},
though still restricted to datatypes that can be turned into groups.
% DBToaster relies on iterated differentiation to

% Like
\paragraph{Logical relations}
To study correctness of incremental programs we use a logical relation among
initial values \ensuremath{\Varid{v}_{1}}, updated values \ensuremath{\Varid{v}_{2}} and changes \ensuremath{\Varid{dv}}.
To define a logical relation for an untyped $\lambda$-calculus we use a
\emph{step-indexed} logical relation,
following~\citep{Appel01,Ahmed2006stepindexed};
in particular, our definitions are closest to the ones by \citet{Acar08}, who
also works with an untyped language, big-step semantics and (a different form
of) incremental computation. However, their logical relation does not mention
changes explicitly, since they do not have first-class status in their system.
Moreover, we use environments rather than substitution, and use a slightly
different step-indexing for our semantics.

% \paragraph{Other work not using change structures}

% in other approaches, a change for an atomic value |a1| simply describes an
% atomic value |a1| replacing it.
% %Self-adjusting computation
% In ILC, we can define changes for |a1 :: A| using a richer language, but the richer
% this language is, the more effort is needed to perform case analysis on it. This
% affects derivatives of primitives handling changes of type |Dt ^ T|.
% % there is a tension between  tradeoff between

% % A precedent is used by
% % \citet{Cicek2016type} to study a type system that describes complexity of
% % self-adjusting computation in the presence of control flow changes during
% % incremental evaluation.

\paragraph{Dynamic incrementalization}
The approaches to incremental computation with the widest applicability are
in the family of self-adjusting computation~\citep{Acar05,Acar09}, including its
descendant Adapton \citep{Hammer2014adapton}.
These approaches incrementalize programs by combining memoization and change
propagation: after creating a trace of base computations, updated inputs are
compared with old ones in $O(1)$ to find corresponding outputs, which are
updated to account for input modifications. Compared to self-adjusting
computation, Adapton only updates results when they are demanded. As usual,
incrementalization is not efficient on arbitrary programs.
To incrementalize efficiently a program must be designed so that input changes
produce small changes to the computation trace; refinement type systems have
been designed to assist in this task~\citep{Cicek2016type,Hammer2016typeda}.
Instead of comparing inputs by pointer equality, Nominal
Adapton~\citep{Hammer2015} introduces first-class labels to identify matching
inputs, enabling reuse in more situations.

Recording traces has often significant overheads in both space and time
(slowdowns of ~20-30$\times$ are common), though
\citet{Acar10TDT} alleviate that by with datatype-specific support for tracing
higher-level operations, while \citet{Chen2014} reduce that overhead by
optimizing traces to not record redundant entries, and by logging chunks of
operations at once, which reduces memory overhead but also potential speedups.
